Case for Support

### 2021-11-17 application/Je-S_sections.tex

The main beneficiaries of this proposal will be experimental brain and animal behaviour labs worldwide who currently lack the highly technical expertise required to apply Machine Intelligence to their experiments. While increasingly automated Machine Learning platforms are becoming available across many data science fields, they require deep expertise or training to construct and train new algorithms, or they are easy to deploy and use but are limited to very specific domains (e.g. face recognition, pose estimation).

The scientific community will benefit from gains in efficiency and reproducibility of deploying Machine Intelligence pipelines and allowing non-experts to compose existing algorithms in a flexible way to design entirely novel Machine Intelligence applications tailor-made for specific research questions. In fields like systems neuroscience, where progress is made essentially by exploring the vast unknown space of animal behaviour and wide-scale brain physiology, the need for new bespoke algorithms and analysis procedures is constant and critical. If successful, this proposal will entirely transform the landscape of Machine Intelligence applied to experimental lab research in biology, by empowering non-technical researchers to recombine and repurpose Machine Intelligence tools using a flexible visual programming environment and providing seamless integration with all the measurement, instrumentation and control packages already available in Bonsai.

In addition to the researchers themselves, the standard analysis pipelines developed in this project will allow anyone who is interested in asking questions about the behaviour and physiology of animals to analyse and understand their data. As an example of these more general applications, increasingly in research institutions worldwide Bonsai is being used to monitor animal welfare by in-house technicians who are concerned about a broader range of environmental variables which might impact animal housing quality but which are beyond the ones which are available in standard vivarium solutions. They use Bonsai as it allows them to prototype and design entirely new monitoring systems on-site, without the need to engage an expensive software development process to answer simple questions. With access to the neuroinformatics packages developed for this proposal, staff at these institutions would be immediately empowered to address more sophisticated questions regarding interactions between housing conditions and animal physiology, which at the moment may be unthinkable to them, but would lead to instantly improving the welfare of hundreds of thousands of research animals in biological research facilities.

>>>>

The main beneficiaries of this proposal will be experimental brain and animal behaviour labs worldwide who currently lack the highly technical expertise required to apply Machine Intelligence to their experiments. In fields like systems neuroscience, where progress is made essentially by exploring the vast unknown space of animal behaviour and wide-scale brain physiology, the need for redesign of experimental protocols and rapid analysis of large and complex datasets is constant and critical. While increasingly automated Machine Learning platforms are becoming available across many data science fields, they require deep expertise or training to construct and train new algorithms, or they are easy to deploy and use but are limited to very specific domains (e.g. face recognition, pose estimation). If successful, this proposal will entirely transform the landscape of Machine Intelligence applied to experimental lab research in biology, by empowering non-technical researchers to recombine and repurpose Machine Intelligence tools using a flexible visual programming environment and providing seamless integration with all the measurement, instrumentation and control packages already available in Bonsai.

Machine learning and computational neuroscience experts will also benefit from a standard platform to compare and benchmark different algorithms and stress test their tools against community datasets and experiments. The addition of new programming languages to develop and interface with Bonsai will also greatly expand the potential for application of their methods into new domains of research.

The scientific community will collectively benefit from gains in efficiency and reproducibility of deploying Machine Intelligence pipelines and allowing non-experts to compose existing algorithms in a flexible way to design entirely novel Machine Intelligence applications tailor-made for specific research questions. The proposed tools will generate cost savings to individual laboratories as well as agencies who fund them by drastically reducing duplicated software development efforts across research groups.

In addition to the research community, the standard analysis pipelines developed in this project may be applied to other domains of society where tracking, monitoring and understanding animal behaviour would yield benefits and improved outcomes. For example, Bonsai is being increasingly used in research institutions worldwide to monitor animal welfare by in-house technicians who are concerned about a broader range of environmental variables which might impact animal husbandry but which are beyond the ones which can be monitored in standard vivarium solutions. As a free and open-source software package, Bonsai would empower such users to prototype and design new animal monitoring systems on-site, without the need to engage in expensive software development to answer simple questions, and would lead to instantly improving the welfare of hundreds of thousands of research animals in biological research facilities.

### 2021-11-17 application/detailsOfResource.tex

Spikes from multiple neurons can be detected in recordings from a single
electrode inserted in neural tissue. Spike sorting is the computational step
used to assign each spike to the neuron that generated it. Most spike sorting
algorithms are designed to work offline (i.e., to use a complete recording
session, after the session terminated). Online spike sorting (i.e., the task
of assigning spikes to individual neurons as recordings are being acquired) is a
still unsolved task, especially for recording devices with large number of
electrodes.
%
However, for the type of BCI application proposed here, previous studies have
shown that spike sorting is not beneficial and high performance can be retained
by just detecting spikes, without assigning them to invidivual
neurons~\citep{trautmannEtAl19,todorovaEtAl14}. Thus, we will only detect spikes with a simple zero-corssing method,
and not assign them to individual neurons.  We will, however, integrate into
Bonsai two offline spike sorting methods, Kilosort~\citep[][written in Matlab]{pachitariuEtAl16}, and MountainSort~\citep[][written in
Python]{chungEtAl17}, to provide Bonsai users with spike sorting functionality for
applications that benefit from it.

@@

    \item[Gaussian Linear Dynamical
        System]\citep[(GLDS)][]{andersonAndMoore12}: with sufficiently large
        bin sizes, spike counts can be modelled as Gaussian random processes.
        This Gaussianity assumption greatly simplifies the estimation of
        parameters of linear dynamical system (LDS) models, as well as
        inferences about their states. After models' parameters have been
        learned, GLDS can be used online. A unique feature of GLDS is that the
        posterior distribution of states given all observation up to the
        present can be calculated efficiently. This estimate of the posterior
        distribution can be used online for experimental control, as we
        propose in Section~\ref{sec:comparisonOfMultipleMethods}. We will use an R implementation of GLDS
        which allows the use of external
        inputs\\ \footnote{https://github.com/joacorapela/kalmanFilter}.

>>>>

Spikes from multiple neurons can be detected in recordings from a single
electrode inserted in neural tissue. Spike sorting is the computational step
used to assign each spike to the neuron that generated it. Most spike sorting
algorithms are designed to work offline (i.e., to use a complete recording
session, after the session terminated). Online spike sorting (i.e., the task
of assigning spikes to individual neurons as recordings are being acquired) is a
still unsolved task, especially for recording devices with large number of
electrodes.
%
However, for the type of BCI application proposed here, previous studies have
shown that spike sorting is not beneficial and high performance can be retained
by just detecting spikes, without assigning them to invidivual
neurons~\citep{trautmannEtAl19,todorovaEtAl14}. Thus, we will only detect spikes with a simple zero-crossing method,
and not assign them to individual neurons.  We will, however, integrate into
Bonsai two offline spike sorting methods, Kilosort~\citep[][written in Matlab]{pachitariuEtAl16}, and MountainSort~\citep[][written in
Python]{chungEtAl17}, to provide Bonsai users with spike sorting functionality for
applications that benefit from it.

@@

    \item[Gaussian Linear Dynamical
        System]\citep[(GLDS)][]{andersonAndMoore12}: with sufficiently large
        bin sizes, spike counts can be modelled as Gaussian random processes.
        This Gaussianity assumption greatly simplifies the estimation of
        parameters of linear dynamical system (LDS) models, as well as
        inferences about their states. After models' parameters have been
        learned, GLDS can be used online. A unique feature of GLDS is that the
        posterior distribution of states given all observation up to the
        present can be calculated efficiently. This estimate of the posterior
        distribution can be used online for experimental control, as we
        propose in Section~\ref{sec:comparisonOfMultipleMethods}. We will use an R implementation of GLDS
        which allows the use of external
        inputs\footnote{https://github.com/joacorapela/ kalmanFilter}.

### 2021-11-17 application/resourceManagementandSustainability.tex

\section{Resource management}
%
As an open source project developed over almost 10 years, Bonsai 
has well-established management practices built around the user community.
%
The source code for the core system and the majority of extension packages
are hosted at \url{github.com/bonsai-rx}.
%
The documentation, including learning materials and all relevant community
links are on a publicly accessible website at \url{bonsai-rx.org}.
%
A community user forum is a public Google Group with more than 700 active users (Figure~\ref{fig:bonsai}f) and several
new questions and answers exchanged every day.
%
Package installation and updates are supported by standalone package
management software included within the core Bonsai distribution and hosted
in a shared centralised and curated repository.
%
This ensures that bugfixes and enhancements are easily available but
also importantly allows for reproducible deployment of user environments
to ensure working experiments are not disrupted by changes in the system.
Large-scale experimental projects such as the International Brain Laboratory
have leveraged this capability of Bonsai to standardise data acquisition and
control software across dozens of laboratories around the world.
%
Community technology previews are made available through the package manager,
allowing early adopters to opt-in to the latest features
before they are generally available. This allows the community
to help report issues with new features on GitHub, thereby
distributing bug reporting and triage and overall improving the quality of
stable releases.

The core compiler and IDE is extensively tested, and takes advantage of
a strong type system to greatly reduce bugs at compile-time. A set of best
practices for leveraging this infrastructure for building new packages is
in place.

@@

NeuroGEARS Ltd is a keystone of this community and a key collaborator in the current proposal.
As a company founded by the original developer of Bonsai, it helps to support and expand the
user base well beyond the life of the current proposal. NeuroGEARS currently provides
quality control and curation for the core Bonsai code, as well as providing training and support
services for research institutions around the world who want to start using Bonsai.
Finally, NeuroGEARS also facilitates knowledge transfer across users and institutions adopting
Bonsai, by moderating the community user forums and collaborating with organisations
who want to build new innovative research projects on top of Bonsai, as in the
current proposal.

The specific contributions developed as part of this project will form new core
components and packages within the Bonsai ecosystem. In particular, the new language integration packages will support long-term sustainability, by dramatically
increasing Bonsai's interoperability with existing software development communities which are
already well established in basic science (MATLAB, Python, R).
%
Developing these new packages with input from the core advisory panel of Bonsai,
and early public distribution and user testing within the SWC and planned workshops, will ensure
solid and broad user engagement from the beginning of the development cycle and will therefore
build momentum for their use, support and dissemination.
%
Close collaboration with the core maintainers of the Bonsai ecosystem at NeuroGEARS will
allow them to take over responsibility for maintenance and support of these packages beyond the
end of the project, following the same principles of open-source and free sharing of knowledge
that characterises the entire Bonsai ecosystem.

>>>>

\section{Resource management}
%
As an open source project developed over almost 10 years, Bonsai 
has well-established management practices built around the user community.
%
The source code for the core system and the majority of extension packages
are hosted at \url{github.com/bonsai-rx}.
%
The documentation, including learning materials and all relevant community
links are publicly accessible at \url{bonsai-rx.org}.
%
A community user forum is a public Google Group with more than 700 active users (Figure~\ref{fig:bonsai}f) exchanging several
new questions and answers every day.
%
Package installation and updates are supported by standalone package
management software included within the core Bonsai distribution and hosted
in a shared centralised and curated repository.
%
This ensures that bugfixes and enhancements are easily available and, importantly, allows for reproducible deployment of user environments
to ensure working experiments are not disrupted by system changes.
Large-scale experimental projects such as the International Brain Laboratory
have leveraged this capability of Bonsai to standardise data acquisition and
control software across dozens of laboratories around the world.
%
Community technology previews are made available through the package manager,
letting early adopters opt-in to the latest features
before they are generally available. This allows the community
to report issues with new features on GitHub, thereby
distributing bug reporting and triage, and improving the quality of
stable releases.

The core compiler and IDE is extensively tested, and uses
a strong type system to greatly reduce bugs at compile-time. Best
practices for leveraging this infrastructure for building new packages are
in place.

@@

NeuroGEARS Ltd is a keystone of this community and a key collaborator in the current proposal.
Founded by the original developer of Bonsai, it helps to support and expand the
user base well beyond the life of the current proposal. NeuroGEARS currently provides
quality control and curation for the core Bonsai code, and provides training and support
services for research institutions around the world who want to start using Bonsai.
NeuroGEARS also facilitates knowledge transfer between users and institutions adopting
Bonsai, by moderating the community user forums and collaborating with organisations
who want to build new innovative research projects on top of Bonsai, as in the
current proposal.

The specific contributions developed as part of this project will form new core
components and packages within the Bonsai ecosystem. In particular, the new language integration packages will support long-term sustainability, by dramatically
increasing Bonsai's interoperability with existing software development communities which are
already well established in basic science (MATLAB, Python, R).
%
Developing these new packages with input from the core advisory panel of Bonsai,
and early public distribution and user testing within the SWC and with planned workshops, will ensure
solid and broad user engagement from the beginning of the development cycle and will therefore
build momentum for their use, support and dissemination.
%
Close collaboration with the core maintainers of the Bonsai ecosystem at NeuroGEARS will
allow them to take over responsibility for maintenance and support of these packages beyond the
end of the project, following the same principles of open-source and free sharing of knowledge
that characterises the entire Bonsai ecosystem.

### 2021-11-17 application/userEngagement.tex

The proposed development will facilitate interfacing with other programming languages
such as Python, R, and MATLAB. This will attract a whole new
community of Bonsai users with expertise in software development. To provide these new users with the full power of Bonsai,
we plan to organise hackathons to accelerate integration of their existing data
analysis and ML algorithms and methods into Bonsai using the new
ML package.

 Given the modular and integrative nature of the ML package, we
 plan to advertise its development early in the project, to allow interested
 partners to contribute to key
 components of the platform, such as the Python and MATLAB integrations. We
 also want to promote the co-development of algorithms by early adopters of the
 package, in true collaborative fashion. Given its modular nature and built-in
 package manager capable of handling dependency resolution and curation of
 community contributed content, Bonsai is well positioned for such an approach.

>>>>

The proposed development will facilitate interfacing with other programming languages
such as Python, R, and MATLAB. This will attract a whole new
community of Bonsai users with expertise in software development. To provide these new users with Bonsai's full power,
we plan to organise hackathons to accelerate integration of their existing data
analysis and ML algorithms and methods into Bonsai using the new
ML package.

 Given the modular and integrative nature of the ML package, we
 plan to advertise its development early in the project, to allow interested
 partners to contribute to key
 components of the platform, such as the Python and MATLAB integrations. We
 also want to promote co-development of algorithms by early adopters, in true collaborative fashion. Given its modular nature and built-in
 package manager capable of handling dependency resolution and curation of
 community contributed content, Bonsai is well positioned for such an approach.
